From Causal Discovery to Intervention Simulation: Modeling Precarity and Depression in the HELIUS Study

### 2025-05-18 draft_v1/draft_v1.qmd
update_draft

Parameter sets with stronger interactions exhibit more pronounced declines in both depression and precarity in response to intervention. At full intervention—modeled as a complete shift of external stress ($S$) toward its observed minimum—the mean depression level decreases from -0.113 to -0.963, and precarity from -0.100 to -0.958. These changes correspond to reductions of approximately 90.8\% relative to their initial magnitudes, indicating that targeted intervention can shift the system toward a substantially lower-risk equilibrium.

@@

[^1]: This resembles a classic *pitchfork bifurcation* in discrete dynamical systems: when the gain $k$ in a scalar map like $x_{t+1} = \tanh(k x_t)$ exceeds a critical value, the system shifts from a single fixed point to two stable attractors and one unstable one. In our model, high values of $\alpha_2 \times \beta_2$ similarly enable the emergence of multiple attractor states.

>>>>

Parameter sets with stronger interaction dynamics exhibited more pronounced reductions in both depression and precariousness following intervention. At full intervention—modeled as shifting the external stressor ($S$) to its observed minimum—the reduction in depression ranged from [0.68*SD* – 0.97*SD*], and for social precariousness from [0.77*SD* – 0.94*SD*], corresponding to the lowest to highest values of parameter strength.[^3] These results suggest that systems with stronger internal dynamics are more responsive to external change, enabling targeted interventions to more effectively shift the system toward a lower-risk equilibrium.

[^3]: Standard deviations (*SD*) are based on the original scale of each variable: $\textit{SD}_D = 1.0$ for depression ($D$) and $\textit{SD}_P = 0.9$ for precariousness ($P$). Parameter strength is defined as the product $\beta_1 \alpha_1 \beta_2 \alpha_2$, reflecting the combined influence of internal feedback and external input sensitivity.

@@

[^1]: This is similar to a well-known pattern in dynamical systems called a *pitchfork bifurcation*. In simple terms, when the system’s internal feedback becomes strong enough—for example, when $k$ in the equation $x_{t+1} = \tanh(k x_t)$ passes a certain threshold—the system changes from having one stable state to having two possible stable states (and one unstable one in between). In our model, this kind of shift happens when the feedback loop between depression and precariousness—captured by the product $\alpha_2 \times \beta_2$—becomes strong. It allows the system to settle into different long-term states, depending on initial conditions.

### 2025-05-18 draft_v1/draft_v1.tex
update_draft

\date{2025-05-14}
\begin{document}
\maketitle
\begin{abstract}
\noindent This study investigates the dynamic relationship between
precarious life conditions and depression by combining constraint-based
causal discovery with computational modeling. Using cross-sectional data
from the HELIUS study, we applied cycle-capable causal discovery
algorithms to uncover directional relationships between financial
stress, domain-specific precariousness indicators, and depressive
symptoms. Our multi-resolution analysis---spanning both aggregate and
symptom-level constructs---identified financial stress as a robust
driver of depression, and highlighted key symptoms, such as sleep
disturbance and depressed mood, as likely initiators or reinforcers of
downstream precariousness. To explore how these structural insights
shape intervention effectiveness, we developed a simplified nonlinear
dynamical model simulating the co-evolution of depression and social
precarity under varying levels of financial stress reduction. Simulation
results revealed that the system's responsiveness depends not only on
the intensity of external intervention but also on internal structural
features such as feedback strength and noise. In particular, some
configurations exhibited bistability, requiring interventions to surpass
a critical threshold to trigger lasting improvement, while others
responded more proportionally. Together, these findings emphasize that
internal feedback dynamics---not just external adversity---can shape the
impact of mental health interventions. More broadly, our study offers a
scalable framework for integrating causal discovery and dynamical
simulation to better understand and inform policy responses to complex
psychosocial phenomena.
\end{abstract}

@@

Parameter sets with stronger interactions exhibit more pronounced
declines in both depression and precarity in response to intervention.
At full intervention---modeled as a complete shift of external stress
(\(S\)) toward its observed minimum---the mean depression level
decreases from -0.113 to -0.963, and precarity from -0.100 to -0.958.
These changes correspond to reductions of approximately 90.8\% relative
to their initial magnitudes, indicating that targeted intervention can
shift the system toward a substantially lower-risk equilibrium.

@@

In our model, bistability arises from the interaction between strong
internal coupling---particularly the product
\(\alpha_2 \times \beta_2\)---and the system's nonlinearity. The
\(\tanh(\cdot)\) activation function saturates at high input values,
meaning sufficient gain is required to push the system into nonlinear
regimes where multiple stable points emerge. If the feedback gain is too
low, or the system is overwhelmed by noise, it behaves more linearly and
settles into a single equilibrium---yielding a unistable
configuration.\footnote{This resembles a classic \emph{pitchfork
  bifurcation} in discrete dynamical systems: when the gain \(k\) in a
  scalar map like \(x_{t+1} = \tanh(k x_t)\) exceeds a critical value,
  the system shifts from a single fixed point to two stable attractors
  and one unstable one. In our model, high values of
  \(\alpha_2 \times \beta_2\) similarly enable the emergence of multiple
  attractor states.}

>>>>

\date{2025-05-18}
\begin{document}
\maketitle
\begin{abstract}
\noindent This study investigates the dynamic relationship between
precarious life conditions and depression by combining constraint-based
causal discovery with computational modeling. Using cross-sectional data
from the HELIUS study, we applied cycle-capable causal discovery
algorithms to uncover directional relationships between financial
stress, domain-specific precariousness indicators, and depressive
symptoms. Our multi-resolution analysis---spanning both aggregate and
symptom-level constructs---identified financial stress as a robust
driver of depression, and highlighted key symptoms, such as sleep
disturbance and depressed mood, as likely initiators or reinforcers of
downstream precariousness. To explore how these structural insights
shape intervention effectiveness, we developed a simplified nonlinear
dynamical model simulating the co-evolution of depression and social
precarity under varying levels of financial stress reduction. Simulation
results revealed that the system's responsiveness depends not only on
the intensity of external intervention but also on internal structural
features such as feedback strength and noise. In particular, some
configurations exhibited bistability, requiring interventions to surpass
a critical threshold to trigger lasting improvement, while others
responded more proportionally. Together, these findings emphasize that
internal feedback dynamics---not just external adversity---can shape the
impact of mental health interventions. More broadly, our study offers a
scalable framework for integrating causal discovery and dynamical
simulation to better understand and inform policy responses to complex
psychosocial phenomena.
\end{abstract}

@@

Parameter sets with stronger interaction dynamics exhibited more
pronounced reductions in both depression and precariousness following
intervention. At full intervention---modeled as shifting the external
stressor (\(S\)) to its observed minimum---the reduction in depression
ranged from {[}0.68\emph{SD} -- 0.97\emph{SD}{]}, and for social
precariousness from {[}0.77\emph{SD} -- 0.94\emph{SD}{]}, corresponding
to the lowest to highest values of parameter strength.\footnote{Standard
  deviations (\emph{SD}) are based on the original scale of each
  variable: \(\textit{SD}_D = 1.0\) for depression (\(D\)) and
  \(\textit{SD}_P = 0.9\) for precariousness (\(P\)). Parameter strength
  is defined as the product \(\beta_1 \alpha_1 \beta_2 \alpha_2\),
  reflecting the combined influence of internal feedback and external
  input sensitivity.} These results suggest that systems with stronger
internal dynamics are more responsive to external change, enabling
targeted interventions to more effectively shift the system toward a
lower-risk equilibrium.

@@

In our model, bistability arises from the interaction between strong
internal coupling---particularly the product
\(\alpha_2 \times \beta_2\)---and the system's nonlinearity. The
\(\tanh(\cdot)\) activation function saturates at high input values,
meaning sufficient gain is required to push the system into nonlinear
regimes where multiple stable points emerge. If the feedback gain is too
low, or the system is overwhelmed by noise, it behaves more linearly and
settles into a single equilibrium---yielding a unistable
configuration.\footnote{This is similar to a well-known pattern in
  dynamical systems called a \emph{pitchfork bifurcation}. In simple
  terms, when the system's internal feedback becomes strong enough---for
  example, when \(k\) in the equation \(x_{t+1} = \tanh(k x_t)\) passes
  a certain threshold---the system changes from having one stable state
  to having two possible stable states (and one unstable one in
  between). In our model, this kind of shift happens when the feedback
  loop between depression and precariousness---captured by the product
  \(\alpha_2 \times \beta_2\)---becomes strong. It allows the system to
  settle into different long-term states, depending on initial
  conditions.}
